---
title: "Chapter 4"
subtitle: "Distribution^[These notes use content from OpenIntro Statistics Slides by Mine Cetinkaya-Rundel.]"
author: |
  | Department of Mathematics & Statistics
  | North Carolina A\&T State University
#date: "8/19/2020"
urlcolor: blue
header-includes:
    \usepackage{multirow}
    \usepackage{graphicx}
    \usepackage{subfigure}
    \usepackage{multicol}
    \usepackage[utf8]{inputenc}
    \usepackage[english]{babel}
    \usepackage{bm}
    \usepackage{amsmath}
    \usepackage{tikz}
    \usepackage{mathtools}
    \usepackage{textcomp}
    \usepackage{fdsymbol}
    \usepackage{siunitx}
output: 
  beamer_presentation:
    fig_caption: true
    latex_engine: xelatex
    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(readr)
library(openintro)
data(COL)
```

# Negative binomial distribution

## Negative binomial distribution

- The **negative binomial distribution** describes the probability of observing the $k^{th}$ success on the $n^{th}$ trial.

- The following four conditions are useful for identifying a negative binomial case:

  \begin{enumerate}
  \item The trials are independent.
  \item Each trial outcome can be classified as a success or failure.
  \item The probability of success $(p)$ is the same for each trial.
  \item The last trial must be a success.
  \end{enumerate}
  
  Note that the first three conditions are common to the binomial distribution
  
P($k^{th}$ success on the $n^{th}$ trial) $= {n-1 \choose k-1}p^k(1-p)^{n-k}$,
Where $p$ is the probability that an individual trial is a success. All trials are assumed to be independent.

## Practice

\alert{A college student working at a psychology lab is asked to recruit 10 couples to participate in a study. She decides to stand outside the student center and ask every $5^{th}$ person leaving the building whether they are in a relationship and, if so, whether they would like to participate in the study with their significant other. Suppose the probability of finding such a person is 10\%. What is the probability that she will need to ask 30 people before she hits her goal?}

## Practice

\alert{A college student working at a psychology lab is asked to recruit 10 couples to participate in a study. She decides to stand outside the student center and ask every $5^{th}$ person leaving the building whether they are in a relationship and, if so, whether they would like to participate in the study with their significant other. Suppose the probability of finding such a person is 10\%. What is the probability that she will need to ask 30 people before she hits her goal?}

Given: $p=0.10,k=10,n=30.$ We are asked to find the probability of $10^{th}$ success on the $30^{th}$ trial, therefore we use the negative binomial distribution.

## Practice

\alert{A college student working at a psychology lab is asked to recruit 10 couples to participate in a study. She decides to stand outside the student center and ask every $5^{th}$ person leaving the building whether they are in a relationship and, if so, whether they would like to participate in the study with their significant other. Suppose the probability of finding such a person is 10\%. What is the probability that she will need to ask 30 people before she hits her goal?}

Given: $p=0.10,k=10,n=30.$ We are asked to find the probability of $10^{th}$ success on the $30^{th}$ trial, therefore we use the negative binomial distribution.

\begin{align*}
P(10^{th} \text{ success on the } 30^{th} \text{ trial}) &= {29 \choose 9} \times 0.10^10 \times 0.90^20
\end{align*}

## Practice

\alert{A college student working at a psychology lab is asked to recruit 10 couples to participate in a study. She decides to stand outside the student center and ask every $5^{th}$ person leaving the building whether they are in a relationship and, if so, whether they would like to participate in the study with their significant other. Suppose the probability of finding such a person is 10\%. What is the probability that she will need to ask 30 people before she hits her goal?}

Given: $p=0.10,k=10,n=30.$ We are asked to find the probability of $10^{th}$ success on the $30^{th}$ trial, therefore we use the negative binomial distribution.

\begin{align*}
P(10^{th} \text{ success on the } 30^{th} \text{ trial}) &= {29 \choose 9} \times 0.10^10 \times 0.90^20 \\
&= 10,015 \times 0.10^10 \times 0.90^20
\end{align*}

## Practice

\alert{A college student working at a psychology lab is asked to recruit 10 couples to participate in a study. She decides to stand outside the student center and ask every $5^{th}$ person leaving the building whether they are in a relationship and, if so, whether they would like to participate in the study with their significant other. Suppose the probability of finding such a person is 10\%. What is the probability that she will need to ask 30 people before she hits her goal?}

Given: $p=0.10,k=10,n=30.$ We are asked to find the probability of $10^{th}$ success on the $30^{th}$ trial, therefore we use the negative binomial distribution.

\begin{align*}
P(10^{th} \text{ success on the } 30^{th} \text{ trial}) &= {29 \choose 9} \times 0.10^10 \times 0.90^20 \\
&= 10,015 \times 0.10^10 \times 0.90^20 \\
&= 0.00012
\end{align*}

## Binomial vs Negative binomial

\alert{How is the negative binomial distribution different from the binomial distribution?}

## Binomial vs Negative binomial

\alert{How is the negative binomial distribution different from the binomial distribution?}

- In the binomial case, we typically have a fixed number of trials and instead consider the number of successes.

- In the negative binomial case, we examine how many trials it takes to observe a fixed number of successes and require that the last observation be a success.

## Practice

\alert{Which of the following describes a case where we would use the negative binomial distribution to calculate the desired probability?}

A) Probability that a 5 year old boy is taller than 42 inches.

B) Probability that 3 out of 10 softball throws are successful.

C) Probability if being dealt a straight flush hand in poker.

D) Probability of missing 8 shots before the first hit

E) Probability of hitting the ball for the $3^{rd}$ time on the $8^{th}$ try.

## Practice

\alert{Which of the following describes a case where we would use the negative binomial distribution to calculate the desired probability?}

A) Probability that a 5 year old boy is taller than 42 inches.

B) Probability that 3 out of 10 softball throws are successful.

C) Probability if being dealt a straight flush hand in poker.

D) Probability of missing 8 shots before the first hit

E) \alert{Probability of hitting the ball for the $3^{rd}$ time on the $8^{th}$ try.}