---
title: "LAB 1 - Intro to R Solution"
author: "MATH224 - Intro to Stat"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(openintro)
```


## Exercise 1

```{r, message=F, warning=F}
(2.59 - 22/7)/(10 - sqrt(23)) #2 Points
```
## Exercise 2

### Part 1

```{r, message=F, warning=F}
log(pi) #2 Points
```

### Part 2

```{r, message=F, warning=F}
exp(2*pi^2) #2 Points
```

OR

```{r, message=F, warning=F}
log(exp(2*pi^2))
```

## Exercise 3

```{r, message=F, warning=F}
a = 3 #4 Points
b = 7

(a+b)^2

a^2 + 2*a*b + b^2
```

As we can see, both $(a+b)^2$ and $a^2+2ab+b^2$ output the same numbers.

## Exercise 4

```{r, message=F, warning=F}
x1 = c(1,2,3,4,5) #4 Points
x2 = c(6,9,1,11,5)

(x1^2 + x2)/2
```

## Exercise 5

```{r, message=F, warning=F}
x3 = c(7,6,8,5,5,9,1) #4 Points
x4 = c(1,2,3)

x3 + x4
```

We still obtain a result even though x3 and x4 are vectors with different lengths. This is because R replicates the smaller vector multiple times to match the size of the larger vector. That is, in the background R creates x4 as `x4 = c(1,2,3,1,2,3,1)` which is of length 7, same as x3. This makes sense based on the output given for the addition.

## Exercise 6

```{r, message=F, warning=F}
glimpse(arbuthnot) #Not necessary for the exercise

arbuthnot$girls #2 Points
```