---
title: "Chapter 6"
subtitle: "Inference for categorical data^[These notes use content from OpenIntro Statistics Slides by Mine Cetinkaya-Rundel.]"
author: |
  | Department of Mathematics & Statistics
  | North Carolina A\&T State University
#date: "8/19/2020"
urlcolor: blue
header-includes:
    \usepackage{multirow}
    \usepackage{graphicx}
    \graphicspath{{images/}}
    \usepackage{subfigure}
    \usepackage{multicol}
    \usepackage[utf8]{inputenc}
    \usepackage[english]{babel}
    \usepackage{bm}
    \usepackage{amsmath}
    \usepackage{tikz}
    \usepackage{mathtools}
    \usepackage{textcomp}
    \usepackage{fdsymbol}
    \usepackage{siunitx}
    \usepackage{xcolor,pifont}
    \usepackage{hyperref}
output: 
  beamer_presentation:
    fig_caption: true
    latex_engine: xelatex
    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(readr)
library(openintro)
library(here)
library(tidyverse)
data(COL)
```


# Chi-square test of independence

## Popular kids

\alert{In the dataset \texttt{popular}, students in grades 4-6 were asked whether good grades, athletic ability, or popularity was most important to them. A two-way table separating the students by grade and by choice of most important factor is shown below. Do these data provide evidence to suggest that goals vary by grade?}

\begin{multicols}{2}

\begin{table}[]
\begin{tabular}{rrrr}
\hline
    & Grades & Popular & Sports \\ \hline
$4^{th}$ & 63     & 31      & 25     \\
$5^{th}$ & 88     & 55      & 33     \\
$6^{th}$ & 96     & 55      & 32     \\ \hline
\end{tabular}
\end{table}

\columnbreak

\includegraphics[width=1\columnwidth]{popular_mosaic.pdf}

\end{multicols}


## Chi-square test of independence

\begin{itemize}
\item The hypotheses are:
\begin{itemize}
\item[$H_0$:] Grade and goals are independent. Goals do not vary by grade.
\item[$H_A$:] Grade and goals are dependent. Goals vary by grade.
\end{itemize}

\pause

\item The test statistic is calculated as
\[ \chi^2_{df} = \sum_{i = 1}^{k} \frac{(O - E)^2}{E} \quad \text{ where } \quad df = (R - 1) \times (C - 1), \]
where $k$ is the number of cells, $R$ is the number of rows, and $C$ is the number of columns.

\noindent\rule{4cm}{0.4pt}

\alert{Note:} We calculate $df$ differently for one-way and two-way tables.

\pause

\item The p-value is the area under the $\chi^2_{df}$ curve, above the calculated test statistic.

\end{itemize}


## Expected counts in two-way tables

**Expected counts in two-way tables**

$Expected \text{ }Count = \frac{(row \text{ }total) \times (column \text{ }total)}{table \text{ }total}$

\pause

\begin{table}[]
\begin{tabular}{rrrr|r}
\hline
    & Grades & Popular & Sports & Total\\ \hline
$4^{th}$ & \textcolor{red}{63}     & \textcolor{blue}{31}      & 25 & 119    \\
$5^{th}$ & 88     & 55      & 33 & 176     \\
$6^{th}$ & 96     & 55      & 32 & 183     \\ \hline
Total	& 247	& 141	& 90	& 478 \\
\end{tabular}
\end{table}


\pause

\textcolor{red}{$E_{row~1, col~1} = \frac{119 \times 247}{478} = 61$}

\pause

\textcolor{blue}{$E_{row~1, col~2} = \frac{119 \times 141}{478} = 35$}

## Expected counts in two-way tables

\alert{What is the expected count for the highlighted cell?}

\begin{table}[]
\begin{tabular}{rrrr|r}
\hline
    & Grades & Popular & Sports & Total\\ \hline
$4^{th}$ & 63     & 31      & 25 & 119    \\
$5^{th}$ & 88     & \textcolor{red}{55}      & 33 & 176     \\
$6^{th}$ & 96     & 55      & 32 & 183     \\ \hline
Total	& 247	& 141	& 90	& 478 \\
\end{tabular}
\end{table}

A) $\frac{176 \times 141}{478}$

B) $\frac{119 \times 141}{478}$

C) $\frac{176 \times 247}{478}$

D) $\frac{176 \times 478}{478}$

## Expected counts in two-way tables

\alert{What is the expected count for the highlighted cell?}

\begin{table}[]
\begin{tabular}{rrrr|r}
\hline
    & Grades & Popular & Sports & Total\\ \hline
$4^{th}$ & 63     & 31      & 25 & 119    \\
$5^{th}$ & 88     & \textcolor{red}{55}      & 33 & 176     \\
$6^{th}$ & 96     & 55      & 32 & 183     \\ \hline
Total	& 247	& 141	& 90	& 478 \\
\end{tabular}
\end{table}

\begin{multicols}{2}

A) \textcolor{red}{$\frac{176 \times 141}{478}$}

B) $\frac{119 \times 141}{478}$

C) $\frac{176 \times 247}{478}$

D) $\frac{176 \times 478}{478}$

\columnbreak

\large\alert{\rightarrow 52}

\normalsize\alert{more than expected \# of 5th graders have a goal of being popular}

\end{multicols}


## Calculating the test statistic in two-way tables

Expected counts are shown in \textcolor{blue}{blue} next to the observed count.

\begin{center}
\begin{tabular}{rrrr|r}
  \hline
 & Grades & Popular & Sports	& Total \\ 
  \hline
$4^{th}$ 	&  63 \textcolor{blue}{61} &  31 \textcolor{blue}{35} &  25 \textcolor{blue}{23}	&119 \\ 
$5^{th}$ 	&  88 \textcolor{blue}{91} &  55 \textcolor{blue}{52} &  33 \textcolor{blue}{33}	& 176 \\ 
$6^{th}$	&  96 \textcolor{blue}{95} &  55 \textcolor{blue}{54} &  32 \textcolor{blue}{34}	& 183 \\ 
   \hline
Total	& 247	& 141	& 90	& 478 \\
\end{tabular}
\end{center}

\vspace{0.5cm}

\pause

\begin{eqnarray*} 
\chi^2 &=& \sum \frac{(63 - 61)^2}{61} + \frac{(31 - 35)^2}{35} + \cdots + \frac{(32 - 34)^2}{34} = 1.3121 \\
\pause
df &=& (R - 1) \times (C - 1) = (3 - 1) \times (3 - 1) = 2 \times 2 = 4 
\end{eqnarray*}


## Calculating the p-value

\alert{Which of the following is the correct p-value for this hypothesis test?}

\centering{\textcolor{red}{$\chi^2=1.3121 \qquad df = 4$}}

\begin{multicols}{2}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
chiTail <- function(df, U, showdf = TRUE, axes = TRUE){
	par(mar=c(2, 1, 1, 1), mgp=c(2.1, 0.8, 0), las=1)
	
	sd <- sqrt(2*df)
	xmax <- max(df + 4.000102*sd,U+2)
	x <- seq(0, xmax, 0.05)
	y <- dchisq(x, df)
	plot(x, y, type='l', axes=FALSE)

	if(axes == TRUE){
		abline(h=0)
		axis(1, at=c(0,U,xmax+3), label = c(0,U,NA), cex.axis = 3)		
	}
	if(axes == FALSE){
		lines(c(0,xmax), c(0,0))
	}
	
	if(showdf == TRUE){
		text(x = 0.8*xmax, y = 0.5*max(y), paste("df =",df), cex = 3)	
	}	
	
	these <- which(x > U)
	X <- x[c(these[1], these, rev(these)[1])]
	Y <- c(0, y[these], 0)
	polygon(X, Y, col='#569BBD')
}

chiTail(df = 4, U = 1.3121)
axis(1, at = 2.5, label = "1.3121", tick = FALSE, cex.axis = 3)
```

\columnbreak

\raggedright
A) More than 0.3

B) Between 0.3 and 0.2

C) Between 0.2 and 0.1

D) Between 0.1 and 0.05

E) Less than 0.001

\end{multicols}


## Calculating the p-value

\alert{Which of the following is the correct p-value for this hypothesis test?}

\centering{\textcolor{red}{$\chi^2=1.3121 \qquad df = 4$}}

\begin{multicols}{2}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
chiTail(df = 4, U = 1.3121)
axis(1, at = 2.5, label = "1.3121", tick = FALSE, cex.axis = 3)
```

\columnbreak

\raggedright
A) **More than 0.3**

B) Between 0.3 and 0.2

C) Between 0.2 and 0.1

D) Between 0.1 and 0.05

E) Less than 0.001

\end{multicols}

## Conclusion

\alert{Do these data provide evidence to suggest that goals vary by grade?}

\begin{itemize}

\item[$H_0$:] Grade and goals are independent. Goals do not vary by grade.

\item[$H_A$:] Grade and goals are dependent. Goals vary by grade. \\

\end{itemize}

\pause

Since p-value is high, we fail to reject $H_0$. The data do not provide convincing evidence that grade and goals are dependent. It doesn't appear that goals vary by grade.
