---
title: "Chapter 8"
subtitle: "Introduction to linear regression^[These notes use content from OpenIntro Statistics Slides by Mine Cetinkaya-Rundel.]"
author: |
  | Department of Mathematics & Statistics
  | North Carolina A\&T State University
#date: "8/19/2020"
urlcolor: blue
header-includes:
    \usepackage{multirow}
    \usepackage{graphicx}
    \graphicspath{{images/}}
    \usepackage{subfigure}
    \usepackage{multicol}
    \usepackage[utf8]{inputenc}
    \usepackage[english]{babel}
    \usepackage{bm}
    \usepackage{amsmath}
    \usepackage{color}
    \usepackage{tikz}
    \usetikzlibrary{shapes}
    \usepackage{mathtools}
    \usepackage{textcomp}
    \usepackage{fdsymbol}
    \usepackage{siunitx}
    \usepackage{xcolor,pifont}
    \usepackage{hyperref}
    \usepackage{mathtools}
    \usepackage[geometry]{ifsym}
output: 
  beamer_presentation:
    fig_caption: true
    latex_engine: xelatex
    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(readr)
library(openintro)
library(here)
library(tidyverse)
library(xtable)
library(broom)
library(faraway)
data(COL)
```

# Line fitting, residuls, and correlation

## Modeling numerical variables

In this unit, we will learn to quantify the relationship between two numerical variables, as well as modeling numerical response variables using a numerical or categorical explanatory variable.

## Powerty vs. HS graduate rate

The **scatterplot** below shows the relationship between HS graduate rate in all 50 US states and DC and the % of residents who live below the poverty line (income below \$23,050 for a family of 4 in 2012).

\begin{columns}

\begin{column}{0.5\textwidth}
```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
poverty = read.table("dataset/poverty.txt", header = T, sep = "\t")

par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 2, cex.axis = 2)
plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2])

```
\end{column}

\begin{column}{0.45\textwidth}

\alert{Response variable?}

\pause

\% in poverty

\pause

\alert{Explanatory variable?}

\pause

\% HS grad

\pause

\alert{Relationship?}

\pause

Linear, negative, moderately strong

\end{column}

\end{columns}

##

The linear model for predicting poverty from high school graduation rate in the US is

\centering{$\hat{\text{poverty}} = 64.78-0.62 \times HS_{grad}$}

\raggedright The "hat" is used to signify that this is an estimate.

##

\alert{The high school graduate rate in Georgia is 85.1\%. What poverty level does the model predict for this state?}

\pause

\centering{$64.78-0.62 \times 85.1 = 12.018$}

## Eyeballing the line

\begin{columns}

\begin{column}{0.35\textwidth}

\alert{Which of the following appears to be the line that best fits the linear relationship between \% in poverty and \% HS grad? Choose one}

\end{column}

\begin{column}{0.6\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5, cex.axis = 1.5)

plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2])
abline(lm(poverty$Poverty ~ poverty$Graduates), col = COL[4], lwd = 3, lty = 1)

y1 = lm(poverty$Poverty ~ poverty$Graduates)$coefficients[1] + 2 + (1.1 * lm(poverty$Poverty ~ poverty$Graduates)$coefficients[2]) * poverty$Graduates
abline(lm(y1 ~ poverty$Graduates), lwd = 3, col = COL[2], lty = 2)

abline(h = 14, lwd = 3, col = COL[5], lty = 3)

y2 = 114 - (12/10) * seq(70,100,1)
abline(lm(y2 ~ seq(70,100,1)), lwd = 3, col = COL[3], lty = 4)

legend("topright", inset = 0.05, c("(a)","(b)","(c)", "(d)"), 
       col = c(COL[4],COL[2],COL[5],COL[3]), lwd = 3, lty = c(1,2,3,4))
```

\end{column}

\end{columns}

## Eyeballing the line

\begin{columns}

\begin{column}{0.35\textwidth}

\alert{Which of the following appears to be the line that best fits the linear relationship between \% in poverty and \% HS grad? Choose one}

\textcolor{blue}{Answer:} \textcolor{blue}{(a) Solid Red Line}

\end{column}

\begin{column}{0.6\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5, cex.axis = 1.5)

plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2])
abline(lm(poverty$Poverty ~ poverty$Graduates), col = COL[4], lwd = 3, lty = 1)

y1 = lm(poverty$Poverty ~ poverty$Graduates)$coefficients[1] + 2 + (1.1 * lm(poverty$Poverty ~ poverty$Graduates)$coefficients[2]) * poverty$Graduates
abline(lm(y1 ~ poverty$Graduates), lwd = 3, col = COL[2], lty = 2)

abline(h = 14, lwd = 3, col = COL[5], lty = 3)

y2 = 114 - (12/10) * seq(70,100,1)
abline(lm(y2 ~ seq(70,100,1)), lwd = 3, col = COL[3], lty = 4)

legend("topright", inset = 0.05, c("(a)","(b)","(c)", "(d)"), 
       col = c(COL[4],COL[2],COL[5],COL[3]), lwd = 3, lty = c(1,2,3,4))
```

\end{column}

\end{columns}

## Residuals

**Residuals** are the leftovers from the model fit: Data = Fit + Residual

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5, cex.axis = 1.5)

plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2], cex.lab = 2, cex.axis = 2)
lm_pov_grad = lm(poverty$Poverty ~ poverty$Graduates)
pred = predict(lm_pov_grad)
x = poverty$Graduates
for(i in 1:length(pred)){
  lines(c(x[i],x[i]), c(poverty$Poverty[i],pred[i]), col = COL[2])
}
abline(lm_pov_grad, col = COL[4], lwd = 3)
```

## Residuals

Residual is the difference between the observed ($y_i$) and predicted ($\hat{y_i}$)

\centering{$e_i = y_i - \hat{y_i}$}

\begin{columns}

\begin{column}{0.5\textwidth}
```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 1.5, cex.axis = 1.5)

plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2], cex.axis = 2, cex.lab = 2)
lm_pov_grad = lm(poverty$Poverty ~ poverty$Graduates)

pred = predict(lm_pov_grad)
x = poverty$Graduates

for(i in c(9,40)){
  lines(c(x[i],x[i]), c(poverty$Poverty[i],pred[i]), col = COL[2])
  text(x[i]+0.5, poverty$Poverty[i], "y", cex = 1.5, col = "blue")
  text(x[i]+1.2, mean(c(poverty$Poverty[i],pred[i])), as.character(round(poverty$Poverty[i] - pred[i],2)), cex = 3, col = "orange")
  text(x[i]-0.5, pred[i], expression(hat(y)), cex = 2, col = COL[4])
}
text(x[9], poverty$Poverty[9] + 0.5, "DC", col = COL[2],cex = 1.8)
text(x[40], poverty$Poverty[40] - 0.5, "RI", col = COL[2],cex = 1.8)

abline(lm_pov_grad, col = COL[4], lwd = 3)
```
\end{column}

\begin{column}{0.45\textwidth}

\pause

\begin{itemize}

\item \% living in poverty in DC is 5.44\% more than predicted.

\pause

\item \% living in poverty in RI is 4.16\% less than predicted.

\end{itemize}

\end{column}

\end{columns}

## Quantifying the relationship

- **Correlation** describes the strength of the \underline{linear} association between two variables.

\pause

- It takes values between -1 (perfect negative) and +1 (perfect positive).

\pause

- A value of 0 indicates no linear association.

## Guessting the correlation

\alert{Which of the following is the best guess for the correlation between \% in poverty and \% HS grad?}

\begin{columns}

\begin{column}{0.3\textwidth}

A) 0.6

B) -0.75

C) -0.1

D) 0.02

C) -1.5

\end{column}

\begin{column}{0.65\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 2, cex.axis = 2)
plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2])
lm_pov_grad = lm(poverty$Poverty ~ poverty$Graduates)
abline(lm_pov_grad, col = COL[4], lwd = 3)
```

\end{column}

\end{columns}

## Guessting the correlation

\alert{Which of the following is the best guess for the correlation between \% in poverty and \% HS grad?}

\begin{columns}

\begin{column}{0.3\textwidth}

A) 0.6

B) \textcolor{red}{-0.75}

C) -0.1

D) 0.02

C) -1.5

\end{column}

\begin{column}{0.65\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 2, cex.axis = 2)
plot(poverty$Poverty ~ poverty$Graduates, ylab = "% in poverty", xlab = "% HS grad", pch=19, col=COL[1,2])
lm_pov_grad = lm(poverty$Poverty ~ poverty$Graduates)
abline(lm_pov_grad, col = COL[4], lwd = 3)
```

\end{column}

\end{columns}


## Guessting the correlation

\alert{Which of the following is the best guess for the correlation between \% in poverty and \% female householder, no husband present?}

\begin{columns}

\begin{column}{0.3\textwidth}

A) 0.1

B) -0.6

C) -0.4

D) 0.9

C) 0.5

\end{column}

\begin{column}{0.65\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 2, cex.axis = 2)
plot(poverty$Poverty ~ poverty$PercentFemaleHouseholderNoHusbandPresent, ylab = "% in poverty", xlab = "% female householder, no husband present", pch=19, col=COL[1,2])

```

\end{column}

\end{columns}

## Guessting the correlation

\alert{Which of the following is the best guess for the correlation between \% in poverty and \% female householder, no husband present?}

\begin{columns}

\begin{column}{0.3\textwidth}

A) 0.1

B) -0.6

C) -0.4

D) 0.9

C) \textcolor{red}{0.5}

\end{column}

\begin{column}{0.65\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
par(mar=c(4,4,1,1), las=1, mgp=c(2.5,0.7,0), cex.lab = 2, cex.axis = 2)
plot(poverty$Poverty ~ poverty$PercentFemaleHouseholderNoHusbandPresent, ylab = "% in poverty", xlab = "% female householder, no husband present", pch=19, col=COL[1,2])

```

\end{column}

\end{columns}

## Assessing the correlation

\alert{Which of the following has the strongest correlation, i.e. correlation coeeficient closest to +1 or -1?}

\begin{columns}

\begin{column}{0.65\textwidth}

```{r, echo=F, message=F, warning=F, out.width="100%",fig.align='center'}
set.seed(179)

# 4 scatterplots

x = seq(0,10,0.1)
yNonLin = (x-3)^2 - 4 + rnorm(length(x), mean = 0, sd = 1)
yPosLinStrong = 3*x + 10 + rnorm(length(x), mean = 0, sd = 2)
yPosLinWeak = 3*x + 10 + rnorm(length(x), mean = 0, sd = 20)
yNegLinWeak = -3*x + 10 + rnorm(length(x), mean = 0, sd = 5)

par(mar=c(2,1,1,1), las=1, mgp=c(0.5,0.7,0), mfrow = c(2,2), cex.lab = 2, cex.axis = 2)

plot(yNonLin ~ x, ylab = "", xlab = "(a)", pch=19, col=COL[1,2], axes = FALSE)
box()

plot(yPosLinStrong ~ x, ylab = "", xlab = "(b)", pch=19, col=COL[1,2], axes = FALSE)
box()

plot(yPosLinWeak ~ x, ylab = "", xlab = "(c)", pch=19, col=COL[1,2], axes = FALSE)
box()

plot(yNegLinWeak ~ x, ylab = "", xlab = "(d)", pch=19, col=COL[1,2], axes = FALSE)
box()
```

\end{column}

\begin{column}{0.3\textwidth}

\pause

\alert{(b) \rightarrow correlation means \underline{linear} association}

\end{column}

\end{columns}
