---
title: "Chapter 4"
subtitle: "Distribution^[These notes use content from OpenIntro Statistics Slides by Mine Cetinkaya-Rundel.]"
author: |
  | Department of Mathematics & Statistics
  | North Carolina A\&T State University
#date: "8/19/2020"
urlcolor: blue
header-includes:
    \usepackage{multirow}
    \usepackage{graphicx}
    \graphicspath{{images/}}
    \usepackage{subfigure}
    \usepackage{multicol}
    \usepackage[utf8]{inputenc}
    \usepackage[english]{babel}
    \usepackage{bm}
    \usepackage{amsmath}
    \usepackage{tikz}
    \usepackage{mathtools}
    \usepackage{textcomp}
    \usepackage{fdsymbol}
    \usepackage{siunitx}
output: 
  beamer_presentation:
    fig_caption: true
    latex_engine: xelatex
    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(readr)
library(openintro)
data(COL)
```


# Geometric distribution

## Milgram experiment

\begin{multicols}{2}

\small
\begin{itemize}

\item Stanley Milgram, a Yale University psychologist, conducted a series of experiments on obedience to authority starting in 1963.

\item Experimenter(E) orders the teacher (T), the subject of the experiment, to give severe electric shocks to a learner (L) each time the learner answers a question incorrectly.

\item The learner is actually an actor, and the electric shocks are not real, but a prerecorded sound is played each time the teacher administers an electric shock.

\end{itemize}
\normalsize

\columnbreak

\includegraphics[width=1\columnwidth]{milgram.png}

\end{multicols}

## Milgram experiment

- These experiments measured the willingness of study participants to obey an authority figure who instructed them to perform acts that conflicted with their personal conscience.

- Milgram found that about 65% of people would obey authority and give such shocks.

- Over the years, additional research suggested this number is approximately consistent across communities and time.

## Bernouilli random variables

- Each person in Milgram's experiment can be thought of as a **trial**.

- A person is labeled a **success** if she refuses to administer a severe shock, and **failure** if she administers such shock.

- Since only 35% of people refused to administer a shock, **probability of success** is **p = 0.35**.

- When an individual trial has only two possible outcomes, it is called a **Bernoulli random variable**.

## Geometric distribution

\alert{Dr. Smith wants to repeat Milgram's experiments but she only wants to sample people until she finds someone who will not inflict a severe shock. What is the probability that she stops after the first person?}

\centering{P($1^{st}$ person refuses) = 0.35}

## Geometric distribution

\alert{Dr. Smith wants to repeat Milgram's experiments but she only wants to sample people until she finds someone who will not inflict a severe shock. What is the probability that she stops after the first person?}

\centering{P($1^{st}$ person refuses) = 0.35}

\raggedright

\alert{... the third person?}

\centering{P($1^{st}$ and $2^{nd}$ shock, $3^{rd}$ refuses) \\ = $\frac{S}{0.65} \times \frac{S}{0.65} \times \frac{R}{0.35} = 0.65^2 \times 0.35 \approx 0.15$}

## Geometric distribution

\alert{Dr. Smith wants to repeat Milgram's experiments but she only wants to sample people until she finds someone who will not inflict a severe shock. What is the probability that she stops after the first person?}

\centering{P($1^{st}$ person refuses) = 0.35}

\raggedright

\alert{... the third person?}

\centering{P($1^{st}$ and $2^{nd}$ shock, $3^{rd}$ refuses) \\ = $\frac{S}{0.65} \times \frac{S}{0.65} \times \frac{R}{0.35} = 0.65^2 \times 0.35 \approx 0.15$}

\raggedright

\alert{... the tenth person?}

## Geometric distribution

\alert{Dr. Smith wants to repeat Milgram's experiments but she only wants to sample people until she finds someone who will not inflict a severe shock. What is the probability that she stops after the first person?}

\centering{P($1^{st}$ person refuses) = 0.35}

\raggedright

\alert{... the third person?}

\centering{P($1^{st}$ and $2^{nd}$ shock, $3^{rd}$ refuses) \\ = $\frac{S}{0.65} \times \frac{S}{0.65} \times \frac{R}{0.35} = 0.65^2 \times 0.35 \approx 0.15$}

\raggedright

\alert{... the tenth person?}

\centering{P($1$ shocks, $10^{th}$ refuses) \\  = $\underbrace{\frac{S}{0.65} \dots \frac{S}{0.65}} \times \frac{R}{0.35} = 0.65^9 \times 0.35 \approx 0.0072$}

## Geometric distribution

**Geometric distribution** describes the waiting time until a success for **independent and identically distributed (iid)** Bernoulli random variables.

- Independence: Outcomes of trials don't affect each other.
- Identical: The probability of success is the same for each trial.

## Geometric distribution

**Geometric distribution** describes the waiting time until a success for **independent and identically distributed (iid)** Bernoulli random variables.

- Independence: Outcomes of trials don't affect each other.
- Identical: The probability of success is the same for each trial.

**Geometric probabilities**

If $p$ represents probability of success, $(1-p)$ represents probability of failure, and $n$ represents number of independent trials

\centering{P(success on the $n^{th}$ trial) = $(1-p)^{n-1}p$}

## Practice

\alert{Can we calculate the probability of rolling a 6 for the first time on the $6^{th}$ roll of a die using the geometric distribution? Note that what was a success (rolling a 6) and what was a failure (not rolling a 6) are clearly defined and one or the other must happen for each trial.}

A) No, on the the roll of a die there are more than 2 possible outcomes.
B) Yes, why not?

## Practice

\alert{Can we calculate the probability of rolling a 6 for the first time on the $6^{th}$ roll of a die using the geometric distribution? Note that what was a success (rolling a 6) and what was a failure (not rolling a 6) are clearly defined and one or the other must happen for each trial.}

A) No, on the the roll of a die there are more than 2 possible outcomes.
B) **Yes, why not?**

\centering{P(6 on the $6^{th}$ roll) = $(\frac{5}{6})^5 \cdot (\frac{1}{6}) \approx 0.067$}

## Expected value

\alert{How many people is Dr. Smith expected to test before finding the first one that refuses to administer the shock?}

## Expected value

\alert{How many people is Dr. Smith expected to test before finding the first one that refuses to administer the shock?}

The expected value, or the mean, of a geometric distribution is defined as $\frac{1}{p}$.

\centering{$\mu = \frac{1}{p} = \frac{1}{0.35} = 2.86$}

## Expected value

\alert{How many people is Dr. Smith expected to test before finding the first one that refuses to administer the shock?}

The expected value, or the mean, of a geometric distribution is defined as $\frac{1}{p}$.

\centering{$\mu = \frac{1}{p} = \frac{1}{0.35} = 2.86$}

\raggedright

She is expected to test 2.86 people before finding the first one that refuses to administer the shock.

## Expected value

\alert{How many people is Dr. Smith expected to test before finding the first one that refuses to administer the shock?}

The expected value, or the mean, of a geometric distribution is defined as $\frac{1}{p}$.

\centering{$\mu = \frac{1}{p} = \frac{1}{0.35} = 2.86$}

\raggedright

She is expected to test 2.86 people before finding the first one that refuses to administer the shock.

But how can she test a non-whole number of people?

## Expected value and its variability

Mean and standard deviation of geometric distribution

\centering{$\mu=\frac{1}{p}$ \hspace{1.5cm}  $\sigma = \sqrt{\frac{1-p}{p^2}}$}

## Expected value and its variability

Mean and standard deviation of geometric distribution

\centering{$\mu=\frac{1}{p}$ \hspace{1.5cm}  $\sigma = \sqrt{\frac{1-p}{p^2}}$}

- Going back to Dr. Smith's experiment:

\centering{$\sigma = \sqrt{\frac{1-p}{p^2}}=\sqrt{\frac{1-0.35}{0.35^2}}=2.3$}

## Expected value and its variability

Mean and standard deviation of geometric distribution

\centering{$\mu=\frac{1}{p}$ \hspace{1.5cm}  $\sigma = \sqrt{\frac{1-p}{p^2}}$}

- Going back to Dr. Smith's experiment:

\centering{$\sigma = \sqrt{\frac{1-p}{p^2}}=\sqrt{\frac{1-0.35}{0.35^2}}=2.3$}

- Dr. Smith is expected to test 2.86 people before finding the first one that refuses to administer the shock, give or take 2.3 people.

## Expected value and its variability

Mean and standard deviation of geometric distribution

\centering{$\mu=\frac{1}{p}$ \hspace{1.5cm}  $\sigma = \sqrt{\frac{1-p}{p^2}}$}

- Going back to Dr. Smith's experiment:

\centering{$\sigma = \sqrt{\frac{1-p}{p^2}}=\sqrt{\frac{1-0.35}{0.35^2}}=2.3$}

- Dr. Smith is expected to test 2.86 people before finding the first one that refuses to administer the shock, give or take 2.3 people.

- These values only make sense in the context of repeating the experiment many many times.
