---
title: "Inference for categorical data Solution"
author: "MATH224 - Intro to Stat"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(openintro)
library(infer)
```

## Exercise 1 (2 Points)

```{r, warning=F, message=F}
yrbss %>%
  count(text_while_driving_30d)
```


## Exercise 2 (3 Points)

### For Section 001

```{r, warning=F, message=F}
yrbss %>%
  select(helmet_12m, text_while_driving_30d)%>%
  na.omit()%>%
  filter(text_while_driving_30d != "did not drive",
         helmet_12m == "never")%>%
  count(text_while_driving_30d)%>%
  mutate(prop = n/sum(n))
```

### For Section 007

```{r, warning=F, message=F}
no_helmet <- yrbss %>%
  filter(helmet_12m == "never")

no_helmet <- no_helmet %>%
  mutate(text_ind = ifelse(text_while_driving_30d == "30", "yes", "no"))%>%
  filter(!is.na(text_ind))

no_helmet%>%
  count(text_ind)%>%
  mutate(p = n/sum(n))
```


## Exercise 3 (3 Points)

```{r, warning=F, message=F}
no_helmet <- yrbss %>%
  filter(helmet_12m == "never")

no_helmet <- no_helmet %>%
  mutate(text_ind = ifelse(text_while_driving_30d == "30", "yes", "no"))%>%
  filter(!is.na(text_ind))

prop_test(no_helmet,
          text_ind ~ NULL,
          success = "yes",
          z = TRUE,
          conf_int = TRUE,
          conf_level = 0.95, correct = FALSE) # 2 Point

(0.07770443 - 0.06519769)/2 # 1 Point
```


## Exercise 4 (3 Points)

```{r, warning=F, message=F}
dat = yrbss%>%
  select(gender)%>%
  na.omit() # 1 Point

prop_test(dat,
          gender ~ NULL,
          success = "male",
          z = TRUE,
          conf_int = TRUE,
          conf_level = 0.95, correct = FALSE) # 1 Point

(0.5205266 - 0.5037094)/2 # 1 Point
```


## Exercise 5 (3 Points)

**2 Point** When p = 0 & 1, ME = 0, but as p increases ME increases. ME maximizes when p = 0.5, after that it starts decreasing as p increases.

```{r, warning=F, message=F}
n <- 1000 #0.25

p <- seq(from = 0, to = 1, by = 0.01) # 0.25 Point
me <- 2 * sqrt(p * (1 - p)/n) # 0.25 Point

dd <- data.frame(p = p, me = me)
ggplot(data = dd, aes(x = p, y = me)) + 
  geom_line() +
  labs(x = "Population Proportion", y = "Margin of Error") # 0.25 Point
```


## Exercise 6 (3 Points)

**2 Points** We would reject the Null Hypothesis $H_0$ because p-values is less than 0.05

```{r, warning = FALSE, message = FALSE}
prop_test(no_helmet,
          text_ind~NULL,
          success = "yes",
          p = 0.08,
          z = TRUE) # 1 Point
```


## Exercise 7 (3 Points)

**1 Point for question** Hypothesis testing on the gender variable to see if the male proportion is 50% or not

```{r, warning = FALSE, message = FALSE}
prop_test(dat,
          gender~NULL,
          success = "male",
          p = 0.50,
          z = TRUE) # 1 Point
```

**1 Point for inference** Since the p-value is less than 0.05, we will reject the null hypothesis of $H_0: p = 0.5$
