---
title: "Chapter 8"
subtitle: "Introduction to Linear Regression"
output: 
  ioslides_presentation:
    progressive: FALSE
    smaller: yes
    widescreen: yes
    transition: "faster"

runtime: shiny_prerendered
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(datasets)
library(tidyverse)
library(shiny)
library(here)
library(faraway)
library(xtable)
library(broom)
library(scales)
library(jpeg)
library(openintro)
library(learnr)
library(readr)
library(knitr)
library(png)
library(gradethis) #remotes::install_github("rstudio/gradethis")
library(learnrhash) #devtools::install_github("rundel/learnrhash")
library(grid)
library(GGally)
library(tinytex)
data(COL)
```


## Acknowledgement

<center>
<div class="black">
These notes use content from OpenIntro Statistics Slides by   

Mine Cetinkaya-Rundel.
</div>
</center>

# Inference for linear regression

## Nature or nurture?

<div class="black">
In 1966 Cyril Burt published a paper called "The genetic determination of differences in intelligence: A study of monozygotic twins reared together and apart". The data consist of IQ scores for [an assumed random sample of] 27 identical twins, one raised by foster parents, the other by the biological parents.

```{r, echo=F, message=F, warning=F, fig.width=6, fig.height=3,fig.align='center'}
data(twins)

r = cor(twins$Foster, twins$Biological)
m1 = lm(twins$Foster ~ twins$Biological)

par(mar=c(4.5,4.5,0.5,0.5))
plot(twins$Foster ~ twins$Biological, xlab = "biological IQ", ylab = "foster IQ", pch = 20, col = COL[1,2], cex = 2, las = 1, cex.axis = 1.5, cex.lab = 1.5)
abline(m1, col = COL[4], lwd = 2)
text(paste("R = ", round(r,3)), x = 75, y = 125, cex = 1.5)
```
</div>




##
<div class="black">
<div class="red">
Which of the following is $\underline{false}$?
</div>

```{r,out.height="30%", out.height="30%"}
summary(m1)
```
\
A) Additional 10 points in the biological twin's IQ is associated with additional 9 points in the foster twin's IQ, on average.
\
B) Roughly 78% of the foster twins' IQs can be accurately predicted by the model.
\
C) The linear model is $\widehat{fosterIQ} = 9.2 + 0.9 \times bioIQ$.
\
D) Foster twins with IQs higher than average IQs tend to have biological twins with higher than average IQs as well.

</div>


##
<div class="black">
<div class="red">
Which of the following is $\underline{false}$?
</div>

```{r, out.height="30%", out.width="30%"}
summary(m1)
```
\
A) Additional 10 points in the biological twin's IQ is associated with additional 9 points in the foster twin's IQ, on average.
\
B)<div class="red">Roughly 78\% of the foster twins'IQs can be accurately predicted by the model.
</div>
\
C) The linear model is $\widehat{fosterIQ} = 9.2 + 0.9 \times bioIQ$.
\
D) Foster twins with IQs higher than average IQs tend to have biological twins with higher than average IQs as well.
</div>



## Testing for the slope
<div class="black">
<div class="red">
Assuming that these 27 twins comprise a representative sample of all twins separated at birth, we would like to test if these data provide convincing evidence that the IQ of the biological twin is a significant predictor of IQ of the foster twin. What are the appropriate hypotheses?
</div>


- $H_0: b_0 = 0; H_A: b_0 \ne 0$ 

- $H_0: \beta_0 = 0; H_A: \beta_0 \ne 0$ 

- $H_0: b_1 = 0; H_A: b_1 \ne 0$ 

- $H_0: \beta_1 = 0; H_A: \beta_1 \ne 0$

</div>


## Testing for the slope
<div class="black">
<div class="red">
Assuming that these 27 twins comprise a representative sample of all twins separated at birth, we would like to test if these data provide convincing evidence that the IQ of the biological twin is a significant predictor of IQ of the foster twin. What are the appropriate hypotheses?
</div>


- $H_0: b_0 = 0; H_A: b_0 \ne 0$ 

- $H_0: \beta_0 = 0; H_A: \beta_0 \ne 0$ 

- $H_0: b_1 = 0; H_A: b_1 \ne 0$ 

- \textcolor{red}{$H_0: \beta_1 = 0; H_A: \beta_1 \ne 0$}

</div>



## Testing for the slope
<div class="black">
$H_0: \beta_1 = 0; H_A: \beta_1 \ne 0$

$$
\begin{eqnarray*}
\hline
Estimate && Std. Error && t value && Pr(>|t|) \\
\hline
\text{(Intercept)} && 9.2076 && 9.2999 && 0.99 && 0.3316 \\
\text{bioIQ} && 0.9014 && 0.0963 && 9.36 && 0.0000\\
\hline
\end{eqnarray*}
$$


- We always use a $t$-test in inference for regression. $\:$ 

- $\color{red}{Test \hspace{0.2cm} statistic, T = \frac{point~estimate - null~value}{SE}}$

- Point estimate = $b_1$ is the observed slope.

- $SE_{b_1}$ is the standard error associated with the slope.

- Degrees of freedom associated with the slope is $df = n - 2$, where $n$ is the sample size.


- **We lose 1 degree of freedom for each parameter we estimate, and in simple linear regression we estimate 2 parameters}, $\beta_0$ and $\beta_1.$**
</div>


## Testing for the slope
<div class= "black">
$H_0: \beta_1 = 0; H_A: \beta_1 \ne 0$

$$
\begin{eqnarray*}
\hline
 & Estimate & Std. Error & t value & Pr(>|t|) \\ 
  \hline
(Intercept) &  9.2076 & 9.2999 & 0.99 & 0.3316 \\ 
  bioIQ & \color{orange}{0.9014}  &   \color{green}{0.0963} & \color{orange}{9.36} & \color{blue}{0.0000} \\ 
   \hline
\end{eqnarray*}
$$
<div>
<div>
$$
\begin{eqnarray*}
T &=& \frac{\color{orange}{0.9014} - 0}{\color{green}{0.0963}} = \color{orange}{9.36} \\
df &=& 27 - 2 = 25 \\
p-value &=& P(|T| > \color{orange}{9.36}) < \color{blue}{0.01}
\end{eqnarray*}
$$
</div>



## % College graduate vs. % Hispanic in LA
<div class="black">
<div class="red">
What can you say about the relationship between \% college graduate and \% Hispanic in a sample of 100 zip code areas in LA?
</div>

<div style="float: left; width: 50%;">

```{r, out.width="100%"}
tyy <- readPNG("tyy.png")
grid.raster(tyy)
```
</div>

<div style="float: right; width: 50%;">
```{r, out.width="100%"}
tpy <- readPNG("tpy.png")
grid.raster(tpy)
```
</div>
</div>



## % College graduate vs. % Hispanic in LA - another look
<div class="black">
<div class="red">
What can you say about the relationship between of \% college graduate and \% Hispanic in a sample of 100 zip code areas in LA?
</div>


```{r, out.width="75%"}
cv <- readPNG("cv).png")
grid.raster(cv)
```
</div>




## % College graduate vs. % Hispanic in LA - linear model
<div class="black">
<div class="red">
Which of the below is the best interpretation of the slope?
</div>

$$
\begin{eqnarray*}
  \hline
 & Estimate & Std. Error & t value & Pr(>|t|) \\ 
  \hline
(Intercept) & 0.7290 & 0.0308 & 23.68 & 0.0000 \\ 
 \%Hispanic & -0.7527 & 0.0501 &-15.01 & 0.0000 \\ 
   \hline
\end{eqnarray*}
$$

- 1% increase in Hispanic residents in a zip code area in LA is associated with a 75% decrease in % of college grads, on average.

- 1% increase in Hispanic residents in a zip code area in LA is associated with a 0.75% decrease in % of college grads, on average.

- An additional 1% of Hispanic residents decreases the % of college graduates in a zip code area in LA by 0.75%.

- In zip code areas with no Hispanic residents, % of college graduates is expected to be 75%.
</div>



## % College graduate vs. % Hispanic in LA - linear model
<div class="black">
<div class="red">
Which of the below is the best interpretation of the slope?
</div>


$$
\begin{eqnarray*}
  \hline
 & Estimate & Std. Error & t value & Pr(>|t|) \\ 
  \hline
(Intercept) & 0.7290 & 0.0308 & 23.68 & 0.0000 \\ 
 \%Hispanic & -0.7527 & 0.0501 & -15.01 & 0.0000 \\ 
   \hline
\end{eqnarray*}
$$


- 1% increase in Hispanic residents in a zip code area in LA is associated with a 75% decrease in % of college grads, on average.

- <div class="red">1% increase in Hispanic residents in a zip code area in LA is associated with a 0.75% decrease in % of college grads, on average.</div>

- An additional 1% of Hispanic residents decreases the % of college graduates in a zip code area in LA by 0.75%.

- In zip code areas with no Hispanic residents, % of college graduates is expected to be 75%.

</div>



## % College graduate vs. % Hispanic in LA - linear model
<div class="black">
<div class=red">
Do these data provide convincing evidence that there is a statistically significant relationship between \% Hispanic and \% college graduates in zip code areas in LA?
</div>

$$
\begin{eqnarray*}
  \hline
 & Estimate & Std. Error & t value & Pr(>|t|) \\ 
  \hline
(Intercept) & 0.7290 & 0.0308 & 23.68 & 0.0000 \\ 
 \%Hispanic & -0.7527 & 0.0501 & -15.01 & 0.0000 \\ 
   \hline
\end{eqnarray*}
$$

Yes, the p-value for % Hispanic is low, indicating that the data provide convincing evidence that the slope parameter is different than 0.

<div class="red">
How reliable is this p-value if these zip code areas are not randomly selected?
</div>

Not very...
</div>



## Confidence interval for the slope
<div class="black">
<div class="red">
Remember that a confidence interval is calculated as $point~estimate \pm ME$ and the degrees of freedom associated with the slope in a simple linear regression is $n - 2$. Which of the below is the correct 95\% confidence interval for the slope parameter? Note that the model is based on observations from 27 twins.
</div>

$$
\begin{eqnarray*}
  \hline
 & Estimate & Std. Error & t value & Pr(>|t|) \\ 
  \hline
(Intercept) & 9.2076 & 9.2999 & 0.99 & 0.3316 \\ 
  bioIQ & 0.9014 & 0.0963 & 9.36 & 0.0000 \\ 
   \hline
\end{eqnarray*}
$$


- $9.2076 \pm 1.65 \times 9.2999$

- $0.9014 \pm 2.06 \times 0.0963$

- $0.9014 \pm 1.96 \times 0.0963$

- $9.2076 \pm 1.96 \times 0.0963$
</div>





## Confidence interval for the slope
<div class="black">
<div class="red">
Remember that a confidence interval is calculated as $point~estimate \pm ME$ and the degrees of freedom associated with the slope in a simple linear regression is $n - 2$. Which of the below is the correct 95\% confidence interval for the slope parameter? Note that the model is based on observations from 27 twins.
</div>


$$
\begin{eqnarray*}
  \hline
 & Estimate & Std. Error & t value & Pr(>|t|) \\ 
  \hline
(Intercept) & 9.2076 & 9.2999 & 0.99 & 0.3316 \\ 
  bioIQ & 0.9014 & 0.0963 & 9.36 & 0.0000 \\ 
   \hline
\end{eqnarray*}
$$
</div>
</div>

## 

<div class="black">

- $9.2076 \pm 1.65 \times 9.2999$

- $\color{red}{0.9014 \pm 2.06 \times 0.0963}$

- $0.9014 \pm 1.96 \times 0.0963$

- $9.2076 \pm 1.96 \times 0.0963$

$$n=27 \qquad df = 27 - 2 = 25 \\95\%:~t^\star_{25} = 2.06 \\0.9014 \pm 2.06 \times 0.0963 \\(0.7 , 1.1)$$
</div>







## Recap
<div class="black">

- Inference for the slope for a single-predictor linear regression model:

- Hypothesis test:
\[ T = \frac{b_1 - null~value}{SE_{b_1}} \qquad df = n - 2 \]

- Confidence interval:
\[ b_1 \pm t^\star_{df = n - 2} SE_{b_1} \]


- The null value is often 0 since we are usually checking for **any** relationship between the explanatory and the response variable.

- The regression output gives $b_1$, $SE_{b_1}$, and **two-tailed** p-value for the $t$-test for the slope where the null value is 0.

- We rarely do inference on the intercept, so we'll be focusing on the estimates and inference for the slope.
</div>



## Caution
<div class="black">

- Always be aware of the type of data you're working with: random sample, non-random sample, or population.

- Statistical inference, and the resulting p-values, are meaningless when you already have population data.

- If you have a sample that is non-random (biased), inference on the results will be unreliable.

- The ultimate goal is to have independent observations.

</div>

