---
title: "Chapter 7"
subtitle: "Inference for numerical data^[These notes use content from OpenIntro Statistics Slides by Mine Cetinkaya-Rundel.]"
author: |
  | Department of Mathematics & Statistics
  | North Carolina A\&T State University
#date: "8/19/2020"
urlcolor: blue
header-includes:
    \usepackage{multirow}
    \usepackage{graphicx}
    \graphicspath{{images/}}
    \usepackage{subfigure}
    \usepackage{multicol}
    \usepackage[utf8]{inputenc}
    \usepackage[english]{babel}
    \usepackage{bm}
    \usepackage{amsmath}
    \usepackage{tikz}
    \usepackage{mathtools}
    \usepackage{textcomp}
    \usepackage{fdsymbol}
    \usepackage{siunitx}
    \usepackage{xcolor,pifont}
    \usepackage{hyperref}
    \usepackage{mathtools}
output: 
  beamer_presentation:
    fig_caption: true
    latex_engine: xelatex
    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(readr)
library(openintro)
library(here)
library(tidyverse)
library(xtable)
data(COL)
```


# One-sample means with the _t_ distribution

## Heights

- According to the CDC, the mean height of U.S. adults ages 20 and older is about 66.5 inches (69.3 inches for males, and 63.8 inches for females).

- In our sample data, we have a sample of 430 college students from a single college.

```{r, echo=F, message=F, warning=F, out.width="75%",fig.align='center'}
set.seed(123)
col_heights = rnorm(n = 430, mean = 67, sd = 5)
hist(col_heights, xlab = "Height (inches)", main = "Histogram of college student heights")
```


## Summary statistics

\small
\begin{center}
\begin{tabular}{ccccc}
  \hline
n & $\bar x$ & s & minimum & maximum \\ 
  \hline
430 & 67.09 & 4.86 & 53.78 & 83.21 \\
   \hline
\end{tabular}
\end{center}

**Objective:** We would like to investigate if the mean height of students at this college is significantly different than 66.5 inches.


## From the Z-Test to the T-Test

Similar to the case of proportions, under certain conditions, we can perform a hypothesis test about the mean $\mu$ using the test statistic

$$Z = \frac{\bar{x}-\mu_0}{\frac{\sigma}{\sqrt{n}}}$$

where $\sigma$ is the population standard deviation and $\mu_0=66.5$ is the hypothesized value for $\mu$.

\pause

- But we do not know $\sigma$ to calculate the SE=$\sigma/\sqrt{n}$

\pause

- We can estimate $\sigma$ using the sample standard deviation $s$.

- The estimated SE will be SE=$s/\sqrt{n}$

\pause

- Then the test statistic becomes

$$T = \frac{\bar{x}-\mu_0}{\frac{s}{\sqrt{n}}}$$

## Conditons

As long as observations are independent, and the population distribution is not extremely skewed, a large sample would ensure that...

  - The sampling distribution of the mean is nearly normal by the central limit theorem.
  
  - The estimate of the standard error, as $\frac{s}{\sqrt{n}}$, is reliable.

## The _t_ distribution

- When the population standard deviation is unknown (almost always), the uncertainty of the standard error estimate is addressed by using a new distribution: the **_t_ distribution**.

\pause

- This distribution also has a bell shape, but its tails are **thicker** than the normal model's.

\pause

- Therefore observations are more likely to fall beyond two SDs from the mean than under the normal distribution.

\pause

- Extra thick tails are helpful for resolving our problem with a less reliable estimate the standard error (since _n_ is small).

```{r, echo=F, message=F, warning=F, fig.width=4, fig.height=1.4,fig.align='center'}
par(mar=c(2, 1, 1, 1), mgp=c(5, 0.6, 0))
plot(c(-4, 4), c(0, dnorm(0)), type='n', axes=FALSE, cex.axis=0.8)
axis(1)
abline(h=0)

X <- seq(-5, 5, 0.01)
Y <- dnorm(X)
lines(X, Y, lty=3, lwd=2)

Y <- dt(X, 2)
lines(X, Y, lwd=1.5, col = "#225588")

legend("topright", inset = 0.05, lty = c(3,1), lwd = c(2,1.5), c("normal","t"), col = c("black","#225588"), bty = "n")
```


## The _t_ distribution

- Always centered at zero, like the standard normal ($z$) distribution.

- Has a single parameter: **degrees of freedom** ($\mathbf{df}$).

```{r, echo=F, message=F, warning=F, fig.width=4, fig.height=1.7,fig.align='center'}
par(mar=c(2, 1, 1, 1), mgp=c(5, 0.6, 0))
plot(c(-3.5, 7), c(0, dnorm(0)), type='n', axes=FALSE)
axis(1, cex.axis = 0.75)
abline(h=0)

COL <- c('#000000FF', '#00000091', '#0000006F', '#0000004B', '#00000022')
DF <- c('normal', 10, 5, 2, 1)

X <- seq(-5, 8, 0.01)
Y <- dnorm(X)
lines(X, Y, col=COL[1])

for(i in 2:5){
	Y <- dt(X, as.numeric(DF[i]))
	lines(X, Y, col=COL[i])
}

legend('topright', legend=c(DF[1], paste('t, df=', DF[2:5], sep='')), col=COL, lty=rep(1, 5), bty = "n", cex = 0.85)

```

\pause

\alert{What happens to shape of the $t$ distribution as $df$ increases?}

\pause

Approaches normal.

## Back to the student heights survey

\small
\begin{center}
\begin{tabular}{ccccc}
  \hline
n & $\bar x$ & s & minimum & maximum \\ 
  \hline
430 & 67.09 & 4.86 & 53.78 & 83.21 \\
   \hline
\end{tabular}
\end{center}

**Objective:** We would like to investigate if the mean height of students at this college is significantly different than 66.5 inches.

## Hypotheses

\alert{What are the hypotheses for testing for the mean of college student heights being different from 66.5 inches?}

\begin{enumerate}[A)]
\item  $H_0: \mu = 66.5$ \\ $H_A: \mu \ne 66.5$
\item  $H_0: \mu = 66.5$ \\ $H_A: \mu > 66.5$
\item  $H_0: \mu = 66.5$ \\ $H_A: \mu < 66.5$ 
\item  $H_0: \mu \ne 66.5$ \\ $H_A: \mu > 66.5$
\end{enumerate}

## Hypotheses

\alert{What are the hypotheses for testing for the mean of college student heights being different from 66.5 inches?}

\begin{enumerate}[A)]
\item  $\textcolor{red}{H_0: \mu = 66.5}$ \\ $\textcolor{red}{H_A: \mu \ne 66.5}$
\item  $H_0: \mu = 66.5$ \\ $H_A: \mu > 66.5$
\item  $H_0: \mu = 66.5$ \\ $H_A: \mu < 66.5$ 
\item  $H_0: \mu \ne 66.5$ \\ $H_A: \mu > 66.5$
\end{enumerate}

## Finding the test statistic

The test statistic for inference on sample mean is the $T$ statistic with $df = n - 1$.

\centering{$T_{df} = \frac{\text{point estimate} - \text{null value}}{SE}$}

\pause

\raggedright
\alert{in context...}


\[\text{point estimate} = \bar{x} = 67.09\]
\pause
\[SE = \frac{s}{\sqrt{n}} = \frac{4.86}{\sqrt{430}} = 0.234 \]
\pause
\[T = \frac{67.09 - 66.5}{0.234} = 2.52\]
\pause
\[df = 430-1 = 429\]

\noindent\rule{4cm}{0.4pt}
\small
\alert{Note:} Null value is 66.5 because in the null hypothesis we set $\mu = 66.5.$


## Finding the p-value

- The p-value is, once again, calculated as the area tail area under the $t$ distribution.

\pause

- Using R:

```{r, echo=T}
2 * pt(2.52, df = 429, lower.tail = FALSE)
```

\pause

- Using a web app:

  [https://gallery.shinyapps.io/dist_calc/](https://gallery.shinyapps.io/dist_calc/)

\pause

- Or when these aren't available, we can use a $t$-table.

## Conclusion of the test

\alert{What is the conclusion of this hypothese test?}

\pause

We saw that the p-value was extremely low. Thus, we reject the null hypothesis. Based on the p-value, we conclude that the survey provide strong evidence that the mean of the college students height is different from the mean height of U.S. adults over 20.

## What is the difference?

- We concluded that there is a difference in the mean heights of the college students compared to the mean height of U.S. adults

\pause

- But it would be more interesting to find out what exactly this difference is.

\pause

- We can use a confidence interval to estimate this difference.

## Confidence interval for a sample mean

- Confidence intervals are always of the form

\centering{$\text{point estimate} \pm ME$}

\pause

- ME is always calculated as the product of a critical value and SE.

\pause

- $ME = t^* \times SE$

\centering{$\text{point estimate} \pm t^* \times SE$}

## Finding the critical $t (t^*)$

- We want to find the 95% confidence interval.

- Using R:

```{r, echo=T}
qt(p = (1+0.95)/2, df = 429)
```

- Or use the $t$-table.

## Constructing a CI for a small sample mean

\alert{Which of the following is the correct calculation of a 95\% confidence interval for the heights of the college students?}

\centering{$\bar{x} = 67.09 \hspace{0.5cm} s = 4.86 \hspace{0.5cm} n = 430 \hspace{0.5cm} SE = 0.234$}

\begin{enumerate}[A)]
\item $66.5 \pm 1.96 \times 0.234$
\item $67.09 \pm 1.97 \times 0.234$
\item $67.09 \pm -2.26 \times 0.234$
\item $66.5 \pm 2.26 \times 4.86$
\end{enumerate}

## Constructing a CI for a small sample mean

\alert{Which of the following is the correct calculation of a 95\% confidence interval for the heights of the college students?}

\centering{$\bar{x} = 67.09 \hspace{0.5cm} s = 4.86 \hspace{0.5cm} n = 430 \hspace{0.5cm} SE = 0.234$}

\begin{enumerate}[A)]
\item $66.5 \pm 1.96 \times 0.234$
\item $\textcolor{red}{67.09 \pm 1.97 \times 0.234 \rightarrow (66.63, 67.55)}$
\item $67.09 \pm -2.26 \times 0.234$
\item $66.5 \pm 2.26 \times 4.86$
\end{enumerate}



## Synthesis

\alert{Does the conclusion from the hypothesis test agree with the findings of the confidence intereval?}

\pause

Yes, the hypothesis test found a significant difference, and the CI does not contain the null value of 66.5.



## Recap: Inference using the $t$-distribution

- If $\sigma$ is unknown, use $t$-distribution with $SE = \frac{s}{\sqrt{n}}$.

\pause

- Conditions:
  
  - Independence of observations (often verified by random sample, and if sampling w/o replacement, n < 10% of population).
  
  - No extreme skew.
  
\pause

- Hypothesis testing:

\centering{$T_{df} = \frac{\text{point estimate}-\text{null value}}{SE}, \text{where }df=n-1$}

\pause

- Confidence interval: $\text{point estimate} \pm t^*_{df} \times SE$

