---
title: "Chapter 4"
subtitle: "Distribution^[These notes use content from OpenIntro Statistics Slides by Mine Cetinkaya-Rundel.]"
author: |
  | Department of Mathematics & Statistics
  | North Carolina A\&T State University
#date: "8/19/2020"
urlcolor: blue
header-includes:
    \usepackage{multirow}
    \usepackage{graphicx}
    \usepackage{subfigure}
    \usepackage{multicol}
    \usepackage[utf8]{inputenc}
    \usepackage[english]{babel}
    \usepackage{bm}
    \usepackage{amsmath}
    \usepackage{tikz}
    \usepackage{mathtools}
    \usepackage{textcomp}
    \usepackage{fdsymbol}
    \usepackage{siunitx}
output: 
  beamer_presentation:
    fig_caption: true
    latex_engine: xelatex
    
    
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r, echo=F, message=F, warning=F}
library(readr)
library(openintro)
data(COL)
```


# Poisson distribution

## Poisson distribution

- The **Poisson distribution** is often useful for estimating the number of rare events in a large population over a short unit of time for a fixed population if the individuals within the population are independent.

- The **rate** for a Poisson distribution is the average number of occurrences in a mostly-fixed population per unit of time, and is typically denotes by $\mathbf{\lambda}$.

- Using the rate, we can describe the probability of observing exactly $k$ rare events in a single unit of time.

P(observe $k$ rare events) = $\frac{\lambda^k e^{-\lambda}}{k!}$,

where $k$ may take a value 0, 1, 2, and so on, and $k!$ represents $k$-factorial. The letter $e \approx 2.718$ is the base of the natural logarithm.

The mean and standard deviation of this distribution are $\lambda$ and $\sqrt{\lambda}$, respectively.

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failures every week. Calculate the probability that in a given week the electricity fails only once.}

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failures every week. Calculate the probability that in a given week the electricity fails only once.}

Given $\lambda = 2.$

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failures every week. Calculate the probability that in a given week the electricity fails only once.}

Given $\lambda = 2.$

\begin{align*}
\text{P(only 1 failure in a week)} &= \frac{2^1 \times e^{-2}}{1!}
\end{align*}

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failures every week. Calculate the probability that in a given week the electricity fails only once.}

Given $\lambda = 2.$

\begin{align*}
\text{P(only 1 failure in a week)} &= \frac{2^1 \times e^{-2}}{1!}\\
&= \frac{2 \times e^{-2}}{1}
\end{align*}

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failures every week. Calculate the probability that in a given week the electricity fails only once.}

Given $\lambda = 2.$

\begin{align*}
\text{P(only 1 failure in a week)} &= \frac{2^1 \times e^{-2}}{1!}\\
&= \frac{2 \times e^{-2}}{1}\\
&= 0.27
\end{align*}

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failure every week. Calculate the probability that on a given \underline{day} the electricity fails three times}


## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failure every week. Calculate the probability that on a given \underline{day} the electricity fails three times}

We are given the weekly failure rate, but to answer this question we need to first calculate the average rate of failure on a given day: $\lambda_{day} = \frac{2}{7} = 0.2857$. Note that we are assuming that the probability of power failure is the same on any day of the week, i.e. we assume independence.

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failure every week. Calculate the probability that on a given \underline{day} the electricity fails three times}

We are given the weekly failure rate, but to answer this question we need to first calculate the average rate of failure on a given day: $\lambda_{day} = \frac{2}{7} = 0.2857$. Note that we are assuming that the probability of power failure is the same on any day of the week, i.e. we assume independence.

\begin{align*}
\text{P(3 failures on a given day)} &= \frac{0.2857^3 \times e^{-0.2857}}{3!}
\end{align*}

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failure every week. Calculate the probability that on a given \underline{day} the electricity fails three times}

We are given the weekly failure rate, but to answer this question we need to first calculate the average rate of failure on a given day: $\lambda_{day} = \frac{2}{7} = 0.2857$. Note that we are assuming that the probability of power failure is the same on any day of the week, i.e. we assume independence.

\begin{align*}
\text{P(3 failures on a given day)} &= \frac{0.2857^3 \times e^{-0.2857}}{3!} \\
&= \frac{0.2857^3 \times e^{-0.2857}}{6}
\end{align*}

## Practice

\alert{Suppose that in a rural region of a developing country electricity power failures occur following a Poisson distribution with an average of 2 failure every week. Calculate the probability that on a given \underline{day} the electricity fails three times}

We are given the weekly failure rate, but to answer this question we need to first calculate the average rate of failure on a given day: $\lambda_{day} = \frac{2}{7} = 0.2857$. Note that we are assuming that the probability of power failure is the same on any day of the week, i.e. we assume independence.

\begin{align*}
\text{P(3 failures on a given day)} &= \frac{0.2857^3 \times e^{-0.2857}}{3!} \\
&= \frac{0.2857^3 \times e^{-0.2857}}{6} \\
&= 0.0029
\end{align*}


## Is it Poisson?

- A random variable may follow a Poisson distribution if the event being considered is rare, the population is large, and the events occur independently of each other.

- However we can think of situations where the events are not really independent. For  example, if we are interested in the probability of a certain number of weddings over one summer, we should take into consideration that weekends are more popular for weddings.

- In this case, a Poisson model may sometimes still be reasonable if we allow it to have a different rate for different times; we could model the rate as higher on weekends that on weekdays.

- The idea of modeling rates for a Poisson distribution against a second variable (day of the week) forms the foundation of some more advanced methods called **generalized linear models**. There are beyond the scope of this course, but we will discuss a foundation of linear models in Chapters 7 and 8.

## Practice

\alert{A random variable that follows which of the following distributions can take on values other than positive integers?}

A) Poisson

B) Negative binomial

C) Binomial

D) Normal

E) Geometric

## Practice

\alert{A random variable that follows which of the following distributions can take on values other than positive integers?}

A) Poisson

B) Negative binomial

C) Binomial

D) \alert{Normal}

E) Geometric

