---
title: "Inference for categorical data"
runtime: shiny
output:
  html_document:
    css: lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, results = FALSE, fig.show = "hide", message = FALSE)
library(tidyverse)
library(openintro)
library(infer)
no_helmet <- yrbss %>%
  filter(helmet_12m == "never")
no_helmet <- no_helmet %>%
  mutate(text_ind = ifelse(text_while_driving_30d == "30", "yes", "no"))
n <- 1000
p <- seq(from = 0, to = 1, by = 0.01)
me <- 2 * sqrt(p * (1 - p)/n)
dd <- data.frame(p = p, me = me)
```


```{r sf-app, echo=FALSE, eval=TRUE, results=TRUE}
inputPanel(
  numericInput("n", label = "Sample size:", value = 300),
  
  sliderInput("p", label = "Population proportion:",
              min = 0, max = 1, value = 0.1, step = 0.01),
  
  numericInput("x_min", label = "Min for x-axis:", value = 0, min = 0, max = 1),
  numericInput("x_max", label = "Max for x-axis:", value = 1, min = 0, max = 1)
)

renderPlot({
  pp <- data.frame(p_hat = rep(0, 5000))
  for(i in 1:5000){
    samp <- sample(c(TRUE, FALSE), input$n, replace = TRUE, 
                   prob = c(input$p, 1 - input$p))
    pp$p_hat[i] <- sum(samp == TRUE) / input$n
  }
  bw <- diff(range(pp$p_hat)) / 30
  ggplot(data = pp, aes(x = p_hat)) +
    geom_histogram(binwidth = bw) +
    xlim(input$x_min, input$x_max) +
    ggtitle(paste0("Distribution of p_hats, drawn from p = ", input$p, ", n = ", input$n))
})
```

Answer the following question with the use of the app above.

1.  Describe the sampling distribution of sample proportions at $n = 300$ and $p = 0.1$.
    Be sure to note the center, spread, and shape.

2.  Keep $n$ constant and change $p$.
    How does the shape, center, and spread of the sampling distribution vary as $p$ changes.
    You might want to adjust min and max for the $x$-axis for a better view of the distribution.

3.  Now also change $n$.
    How does $n$ appear to affect the distribution of $\hat{p}$?
