---
title: "Inference for numerical data Solution"
author: "MATH224 - Intro to Stat"
date: "3/20/2022"
output: pdf_document
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(openintro)
library(infer)
```

## Exercise 1 (5 Points)

**1 Point** H0: p = 0.66, HA: p not = 0.66

OR

$H_0: p = 0.66, H_A: p \ne 0.66$

```{r, warning = FALSE, message = FALSE}
yrbss <- yrbss %>% 
  mutate(physical_3plus = if_else(physically_active_7d > 2, "yes", "no")) #1 Point

yrbss%>%
  filter(!is.na(physical_3plus))%>%
  prop_test(response = physical_3plus,
            success = "yes",
            p = 0.66,
            alternative = "two-sided",
            z = T) #1 Point for hypothesis test

yrbss%>%
  filter(!is.na(physical_3plus))%>%
  prop_test(response = physical_3plus,
            success = "yes",
            alternative = "two-sided",
            z = T,
            conf_int = T,
            conf_level = 0.98) #1 Point for confidence interval
```

**0.5 Points** Since the p-value is greater than 0.02, we fail to reject the null hypothesis.

**0.5 Points** CI: (0.6596, 0.6785)


## Exercise 2 (3 Points)

**1 Point** H0: pmale - pfemale = 0, HA: pmale - pfemale not = 0

OR

$H_0: p_{male} - p_{female} = 0, H_A: p_{male} - p_{female} \ne 0$

```{r, warning = FALSE, message = FALSE}
yrbss%>%
  filter(!is.na(physical_3plus), !is.na(gender))%>%
  prop_test(response = physical_3plus,
            success = "yes",
            explanatory = gender,
            order = c("male","female"), 
            z = TRUE,
            conf_int = TRUE,
            conf_level = 0.95)#1 Point
```


**0.5 Points** Since p is less than 0.05, we reject the null hypothesis.

**0.5 Points** CI: (0.1962, 0.2313)

## Exercise 3 (3 Points)

**1 Point** H0: mu = 66.82, HA: mu not = 66.82

OR

$H_0: \mu = 66.82, H_A: \mu \ne 66.82$


```{r, warning = FALSE, message = FALSE}
yrbss%>%
  filter(!is.na(weight))%>%
  t_test(response = weight,
         mu = 66.82,
         conf_int = TRUE,
         conf_level = 0.95) #1 Point
```

**0.5 Points** Since p-value is less than 0.05, we reject the null hypothesis.

**0.5 Points** CI: (67.61, 68.20)

## Exercise 4 (3 Points)

**2 Points** Explanation: There's seems to be a slight difference in the violin plots. Physically inactive students have a higher standard deviation compared to the physically active students. Physically inactive students have a slightly lower mean weight.

```{r, warning = FALSE, message = FALSE}
yrbss %>%
  filter(!is.na(physical_3plus), !is.na(weight))%>%
  ggplot(aes(x = physical_3plus, y = weight, fill = physical_3plus))+
  geom_violin() #1 Point
```


## Exercise 5 (3 Points)

**2 Points** There is an observable difference. But the difference isn't big enough for us to deem it statistically significant without an hypothesis test.


```{r, warning = FALSE, message = FALSE}
yrbss %>%
  filter(!is.na(physical_3plus), !is.na(weight))%>%
  group_by(physical_3plus) %>%
  summarise(mean_weight = mean(weight)) #1 Point
```


## Exercise 6 (3 Points)

**1 Point** H0: mu_yes - mu_no = 0, HA: mu_yes - mu_no not = 0

OR

$H_0: \mu_{yes} - \mu_{no} = 0, H_A: \mu_{yes} - \mu_{no} \ne 0$

```{r, warning = FALSE, message = FALSE}
yrbss%>%
  filter(!is.na(weight))%>%
  t_test(response = weight,
         explanatory = physical_3plus,
         order = c("yes","no"),
         mu = 0,
         conf_int = TRUE,
         conf_level = 0.95) #1 Point
```

**0.5 Points** Since p-value is less than 0.05, we reject the null hypothesis.

**0.5 Points** CI: (1.12, 2.42)